---
title: "Readme"
output: md_document
---

```{r message=FALSE, warning=FALSE, include=FALSE}
source("rfl-war.R")
```

# Fantasy WAR

[Source](https://www.fantasypoints.com/nfl/articles/season/2021/fantasy-war-part-1-theory#/)

Who was the best fantasy football player of the year? Was it the player who scored the most points by any player at any position? Was it the the early first-round pick who led all RBs? What about the player, who outscored the #2 of his position by over 50 points? You might think there is no way to tell for sure and “to each their own.” Think again.

I’m here to introduce a new statistic called Wins Above Replacement (WAR) for fantasy football, and it directly answers this question. WAR is a fine-tuned tool capable of controlling for everything imaginable that reveals the estimated number of wins each and every player provided the average fantasy team in a given year.

WAR was built to power redraft managers on draft day. When you sit down to make each selection this August, you should have one question on your mind: “Who is going to give my fantasy team the most wins?” That’s it. Nothing else matters. Of course, there’s no way to tell for sure which specific player will provide the most wins for your team. But with WAR, you’ll have the almanac at your hands revealing what the optimal draft looked like for 2020. This in turn will help you make key decisions for your 2021 selections.

In this introduction, a thorough explanation of this model will be explored in the following three parts:

1. Wins Above Replacement Theory
2. Positional Hierarchy
3. Year-Over-Year WAR Trends

## Part 1: Wins Above Replacement Theory

WAR was originally created as a baseball statistic in the 1980s to estimate the number of wins a baseball player provided his team in a given year by controlling for everything else around him. The core concept is the same for fantasy football. WAR will tell fantasy managers how many more head-to-head matchups they should expect to win in a season from starting a particular player vs. that player’s expected replacement-level fill-in.

In order to make this happen, specific information on the following will need to be established in order to convert this baseball sabermetric to the fantasy football world:

1. League Settings
2. Average weekly team score
3. Average team standard deviation
4. “Replacement Level” Definition
5. Player of focus

### #1 League Settings

**Adapted to RFL**

The starting lineup consist of QB, 2 RB, 2 WR, TE, 2 Flex, PK, 2 DL, 2 LB, 2 DB, 3 IDP.

We have 36 Teams in the league, but 3 copies per player. So, we handle the league basically as a 12-team league.

Let's load our base data.

```{r base data, echo=TRUE, message=FALSE, warning=FALSE}
head(starter, 5)
```

### #2: Average Team Weekly Score

**Adapted to RFL**

The average team weekly score is calculated by adding the individual scores of all starting positions together. In order to assign an average score for each starting roster spot, we'll use the average points from the top-x started players of each regular season week.

| Position | Top-x |
|----------|-------|
| QB       | 12    |
| RB       | 24    |
| WR       | 24    |
| TE       | 12    |
| PK       | 12    |
| DL       | 24    |
| LB       | 24    |
| DB       | 24    |

Take QB for example. The league has 12 starting QBs per week and 13 total weeks in a season. This means there are 156 starting QB outputs. The 12 most-started QBs in each week make up these 156 performances, which are simply averaged to provide the expected fantasy points from an average starting QB.

Since we are playing with 3 player copies, the amount of started player can be between 0 (nobody hast started him) and 3 (all three teams started him). In later calculations we include this count to reflect the as starter considered players.

```{r starter for each week, echo=TRUE}
head(eligable_players, 5)
```

This process is repeated for all other positions. Since some of them require two starting slots, the average points from the top-24 started per week is used. The process is the same for flex and IDP as a single starting position, except that all players considered a starter in their native position are not eligible for flex, as it's presumed these players already occupy a starting spot somewhere in the league. However, the next top-24 most-started combination of RB, WR, and TE for offense and DL, LB and DB for defense each week do qualify.

Simply adding all positional averages from this process together produces the expected amount of points the average team in a 12-team half-point PPR league for any given week.

For `r current_season`, that turned out to be `r round(avg_team_scores$points, 2)`.

### #3 Average Team Standard Deviation

The process of determining the average points produced from each individual starting roster spot is performed the same for standard deviation. The only difference is that the standard deviation is calculated on these positions instead of the average. But the process for determining the team-level standard deviation is more complex and requires each individual position’s standard deviation plugged into the following formula:

sd_avg_team = √(∑(sd_i^2))

i = positions * starter

To briefly explain the equation above, each individual starting position’s standard deviation is squared and added together. The square root of the resulting figure provides the team-level standard deviation of the average team. For the average team in `r current_season`, the standard deviation turned out to be `r round(avg_team_scores$sd, 2)`.

In plain terms, this means that about 68% of fantasy team weekly scores fell between `r round(avg_team_scores$sd, 2)` points below or above the league average of `r round(avg_team_scores$points, 2)`. The idea that about 70 percent of weekly team scores in the league were between `r round(avg_team_scores$points - avg_team_scores$sd, 2)` and `r round(avg_team_scores$points + avg_team_scores$sd, 2)` seems about right. And if stretched to two standard deviations, 95 percent of team scores fell between `r round(avg_team_scores$points - (avg_team_scores$sd * 2), 2)` and `r round(avg_team_scores$points + (avg_team_scores$sd * 2), 2)`

Why does knowing team-level standard deviation matter for Wins Above Replacement? In order to ultimately estimate the likelihood that any individual player will lead your team to win a head-to-head matchup, the distribution of possible outcomes expected from your opponent must be known. We’ll come back to this soon.

### #4 Replacement Level Definition

The most critical piece of the WAR puzzle is to determine how valuable a replacement-level player at each position is. But first, we need to figure out what replacement exactly means. The idea of a replacement player is that if you have a starting player miss a game due to injury, suspension, bye, etc., you are forced to insert your next best option into your starting lineup. This would either be someone on your bench or from the waiver wire/free agent pool.

Let’s draw this out with QB. In a 12-team league, WAR assumes the top-12 league started QBs are in one of the twelve team’s lineups.

Let’s say your QB is on bye and you either have to either stream or use a backup. The question is: how many points can you expect out of the “next-best” QB? To figure this out, we need to find the average of the next top-12 started QBs after all determined league starting QBs for each week. We use the top-12 per week because we can’t assume our team will have the best replacement-QB in our league or the worst, so we use the average. All in all, 156 performances (12 per week * 13 weeks) for replacement-level QBs are averaged revealing the estimated points from a replacement-QB.

The process is the same for all positions, with the exception that twice the amount of starters are removed from consideration for RB and WR, due to the fact that each has two starting roster spots. From there, each position has a figure for what is estimated out of a replacement-level player. Here is what each position’s expected weekly points out of a replacement turned out to be in 2020:

```{r replacement data, echo=FALSE, message=FALSE, warning=FALSE}
head(replacement_player_points[1:2], 8)
```

These figures will be used at the end of the next section to compare any player against at his position as the final step in the WAR process.

### #5 Player of Focus

Now comes the fun part. We are ready to use all this information to calculate a player’s WAR. The idea is to figure out the odds the average fantasy team has of winning each matchup when all that is known are the points the player of focus (POF) provided throughout the year.

Example: In Week 1, the POF scored 20 fantasy points, which happened to be `r 20 - avg_player_scores$points_average_player[avg_player_scores$pos=="RB"]` points more than an average starting RB’s `r avg_player_scores$points_average_player[avg_player_scores$pos=="RB"]`. Let's say, our POF is a RB. So the rest of the starting positions on our team are expected to perform to league average as earlier explained. With the average expected points of all other starting roster spots on our team added together along with the POF's 20 points, the POF owner would expect to have scored `r avg_team_scores$points - avg_player_scores$points_average_player[avg_player_scores$pos=="RB"] + 20` points in Week 1 without knowing anything about any other player.

Now we need to know what to expect out of the Kamara owner’s opponent. This is done by adding the average contribution of each position to obtain the expected team output: `r avg_team_scores$points`.

Here’s what this all looks like on paper so far:

#### Week 1 Examples
```{r example week 1, echo=FALSE, message=FALSE, warning=FALSE}
example <- tibble(
  position = c("QB", rep("RB", 2), rep("WR", 2), "TE", rep("FLEX", 2), "PK", rep("DL", 2), rep("LB", 2), rep("DB", 2), "TOTAL"),
  opponent = c(avg_player_scores$points_average_player[avg_player_scores$pos=="QB"], rep(avg_player_scores$points_average_player[avg_player_scores$pos=="RB"], 2), rep(avg_player_scores$points_average_player[avg_player_scores$pos=="WR"], 2), avg_player_scores$points_average_player[avg_player_scores$pos=="TE"], rep(avg_player_scores$points_average_player[avg_player_scores$pos=="FLEX"], 2), avg_player_scores$points_average_player[avg_player_scores$pos=="PK"], rep(avg_player_scores$points_average_player[avg_player_scores$pos=="DL"], 2), rep(avg_player_scores$points_average_player[avg_player_scores$pos=="LB"], 2), rep(avg_player_scores$points_average_player[avg_player_scores$pos=="DB"], 2), avg_team_scores$points),
  "POF Team" = c(avg_player_scores$points_average_player[avg_player_scores$pos=="QB"], avg_player_scores$points_average_player[avg_player_scores$pos=="RB"], 20, rep(avg_player_scores$points_average_player[avg_player_scores$pos=="WR"], 2), avg_player_scores$points_average_player[avg_player_scores$pos=="TE"], rep(avg_player_scores$points_average_player[avg_player_scores$pos=="FLEX"], 2), avg_player_scores$points_average_player[avg_player_scores$pos=="PK"], rep(avg_player_scores$points_average_player[avg_player_scores$pos=="DL"], 2), rep(avg_player_scores$points_average_player[avg_player_scores$pos=="LB"], 2), rep(avg_player_scores$points_average_player[avg_player_scores$pos=="DB"], 2), avg_team_scores$points - avg_player_scores$points_average_player[avg_player_scores$pos=="RB"] + 20)
  
)

example
```
Notice that all expected scores are the same for each position on both teams with the only exception being the POF's 20. This gives the POF owner an expected advantage over the average fantasy opponent.

There’s one problem here. We can’t just assume with 100% certainty that the POF led our team to victory in Week 1 simply due to the higher projected team total. This is where the team-level weekly standard deviation formula comes into play. Maybe our opponent does score exactly `r avg_team_scores$points` points; but maybe they score 300 – or only 180. As any fantasy player knows, it’s going to vary week to week.

To account for this variance, a cumulative distribution function takes the average opponent expected score (`r `r avg_team_scores$points``) and the average team standard deviation (`r `r avg_team_scores$sd`) and determines the likelihood of a score in this presumed normal distribution being below our team’s expected score of `r avg_team_scores$points - avg_player_scores$points_average_player[avg_player_scores$pos=="RB"] + 20`. And for the POF in Week 1, this turned out to be 64%. Simply put: Without knowing the results of any other player than Alvin Kamara in a head-to-head matchup for Week 1, Fantasy players had an estimated 64% chance of winning the matchup after scoring 20 points.

The same process is repeated for all regular season weeks which provides estimated win percentages for all matchups. For all games the POF missed (bye, injury, suspension), the expected output of a replacement-level Player is used to fill in his stead. Once all expected fantasy team win percentages are calculated for the POF owner, the seasonal fantasy team win percentage is simply the average of all weeks.

Now we know the expect win percentage solely due to the contribution of the POV. This is great and all, but we still need to figure out how much better this is than a replacement-level player. So we calculate a win percentage fo rthe replacement level player the same way we did for the POF. Last but not least, we substract the expected wins of the replacement level player from the POF's and the result: WAR.

Since we are playing double header, we multiply the WAR by 2.

And that is it. That is the full process to complete a fantasy football player’s WAR. Everything is simply repeated for every player until all relevant names are crunched and a full list of the most important players emerge.

### Results `r current_season`

```{r result, echo=FALSE, message=FALSE, warning=FALSE}
war
```


